---
title: "Correction for Confounding Factors"
date: "2024-11-21"
author: "Nikolai Köhler"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.dim = c((1+sqrt(5))/2, 1) * 5, cache = TRUE, autodep = TRUE) 
options(width = 70)
```

```{r}
library(ggfortify)
library(tidyverse)
```

For this exercise we are using the `sva` package with gene expression data form a bladder cancer cohort.

```{r}
BiocManager::install("sva")
BiocManager::install("bladderbatch")

library(sva)
library(bladderbatch)
data(bladderdata)

expr <- exprs(bladderEset)
metadata <- pData(bladderEset)
metadata$batch <- as.factor(metadata$batch)
```

# Manual Computation

Generate a PCA and plot with batch and cancer annotation using the `ggfortify` package.
*Hint*: mind the sample/column orientation of `expr`

```{r}
# TODO
```

Before looking at more sophisticated approaches, we do the most simply correction by using the resiudals of a linear model.
Compute them for a `gene ~ cancer` model for each gene.
*Hint*: use the `lm` function and the `residuals` attribute of the `lm` output.

```{r}
# TODO
```

Plot the PCA for all residuals with batch and cancer annotation
```{r}
res_pca <- prcomp(res) 
autoplot(#TODO) + theme_minimal()
```


# Using sva

sva is a package for confounder correction that includes multiple approaches.
The simplest is to compare the outputs of two models, a null model which contains the confounding variables and the actual model that includes confounders and variables of interest.

First generate a null model that includes the batch effect using `model.matrix(<model>, data=metadata)`

```{r}
null_mod <- # TODO
```

Now generate the test model which includes batch and cancer

```{r}
test_mod <- # TODO
```

To compute the p-values, we use an F-test as implemented in the `f.pvalue` function.
It takes in the expression data, test and null model.
Compute the p-values and correct them with the Benjamini-Hochberg method.
```{r}
pvals <- # TODO
qvals <- # TODO
```

# ComBat

ComBat uses an Empirical Bayes approach to correct batch effects (recall Tuesday afternoon).
The `ComBat` function takes in the expression data, a batch vector and a model formula, that does *not* include the batch.
Compute the corrected expression matrix below.

```{r}
combat_mod <- model.matrix(~ 1, data=metadata)
combat_expr = ComBat(# TODO)
```

Plot the PCA for ComBat
```{r}
combat_pca <- prcomp(t(combat_expr))
autoplot(# TODO) + theme_minimal()
```

Now run F-tests on the corrected data to test gene expression between cancer groups.
```{r}
test_mod_combat <- # TODO
pvals_combat <- # TODO
qvals_combat <- # TODO
```

# Comparison

Extract the significant genes from the sva and the ComBat-based p-values.
Compute their overlap by using the `setdiff` function.

```{r}
top_sva <- # TODO
top_combat <- # TODO
setdiff(# TODO)
```

Select the top 50 genes from the combat results and plot a heatmap with the
*uncorrected* expression values.

```{r}
library(pheatmap)

pheatmap(
  expr[top_combat,],
  annotation_col=# TODO
)
```

Now make the same plot with the *corrected* expression values.

```{r}
pheatmap(
  combat_expr[top_combat,],
  annotation_col=# TODO
)
```

Print the session info
```{r}
# TODO
```

## Resources

This exercise is based on the [`sva` vignette](https://bioconductor.org/packages/devel/bioc/vignettes/sva/inst/doc/sva.pdf)